---
title: "Data manipulation"
toc: true
toc-location: right
---

```{r}
#| echo: false
library(tidyverse)
cats <- read_csv("data-raw/cat-coats.csv")
```


# Introduction

-   Summarising data
    -   Whole column
    -   By groups in another column
-   Filtering rows
-   Adding columns
-   Writing data to a file

# Summarising

## Whole column

ðŸŽ¬ Find the mean with:

```{r}
cats |> 
  summarise(mean(mass))
```



### naming the column


ðŸŽ¬ Give a name to the new column containing the mean:

```{r}
cats |> 
  summarise(mean = mean(mass))
```



### adding other summary stats

ðŸŽ¬ Find the mean, standard deviation and sample size:

```{r}
cats |>
  summarise(mean = mean(mass),
            sd = sd(mass),
            n = length(mass))
```




ðŸ› you add se

```{r}
cats |> 
  summarise(mean = mean(mass),
            sd = sd(mass),
            n = length(mass),
            se = sd / sqrt(n))
```

## By groups given in another column

```{r}
cats |> 
  group_by(coat) |> 
  summarise(mean = mean(mass),
            sd = sd(mass),
            n = length(mass),
            se = sd / sqrt(n))
```

## Save the results with assignment

```{r}
cats_summary <- cats |> 
  group_by(coat) |> 
  summarise(mean = mean(mass),
            sd = sd(mass),
            n = length(mass),
            se = sd / sqrt(n))
```
# Filtering rows

## filter to only the black cats

```{r}
cats_black <- cats |> 
  filter(coat == "black")
```


## piping is great

```{r}
cats |> 
  filter(coat == "black") |> 
  summarise(mean = mean(mass),
            n = length(mass))
```

## filter for cats over 4 kilos

```{r}
cats |> 
  filter(mass > 6,
         coat == "ginger")
```


## filter the raw data to remove calico

```{r}
cats |> 
  filter(coat != "calico")
```

```{r}
cats_summary |> 
  filter(coat != "calico")
```

```{r}
cats_summary |> 
  filter(!is.na(sd))
```


```{r}
cats_summary |> 
  drop_na(sd)
```

## Adding columns

```{r}
cats <- cats |> 
  mutate(mass_g = mass * 1000)
```


```{r}
cats <- cats |> 
  mutate(is_ginger_chonk = mass > 6 & coat == "ginger")
```

```{r}
cats <- cats |> 
  mutate(size = case_when(mass < 3.5 ~ "small",
                          mass > 5 ~ "large",
                          .default = "medium"  ))
```


```{r}
cats <- cats |> 
  mutate(size2 = case_when(
    mass < 3.5 ~ "small",
    mass >= 3.5 & mass <= 5 ~ "medium",
    mass > 5 ~ "large"))
```
# Writing data to a file


```{r}
write_csv(cats, "cats-added.csv")
```

Pages made with R [@R-core], Quarto [@Allaire_Quarto_2024], `knitr` 
[@knitr1; @knitr2; @knitr3], `kableExtra` [@kableExtra]

# References